---
title: "Curriculum Vitae"
---

```{r setup, include=FALSE}
source("R/helpers.R")
```

[Download PDF version](cv.pdf){.btn-pill .btn-filled}

## Education

- **Ph.D. Economics**, University of Hyderabad, 2019<br>
  *Dissertation:* Monetary Policy, Uncertainty, Inflation and Output in India<br>
  *Supervisor:* Prof. Bandi Kamaiah
- **MA Economics**, University of Hyderabad, 2010
- **BA (Hons) Economics**, Aligarh Muslim University, 2008

## Academic Appointments

- **Assistant Professor**, Department of Economics, Aligarh Muslim University, Aligarh — 2018–present
- **Assistant Professor**, Department of Economics, MANUU, Hyderabad — 2014–2018

## Research Experience

- **Research Fellow**, School of Economics, University of Hyderabad — 2011–2014
- **Research Associate**, Ministry of Commerce & Industry, Govt. of India — 2010–2011

## Scholarships & Awards

```{r awards, results="asis"}
print_awards()
```

## Workshops & Training

```{r workshops, results="asis"}
print_workshops()
```

## Academic / Administrative Service

```{r service, results="asis"}
print_service()
```

## Academic Affiliation

```{r memberships, results="asis"}
print_memberships()
```

## Skills

```{r skills, results="asis"}
print_skills()
```

## Contact

Department of Economics, Faculty of Social Sciences
Aligarh Muslim University, Aligarh, UP 202002, India
Email: [zulquar.ec@amu.ac.in](mailto:zulquar.ec@amu.ac.in)

*Referee details are available in the [PDF version of the CV](cv.pdf).*
