---
title: "Teaching"
listing:
  - id: course-pages
    contents: courses
    type: default
    sort: "date desc"
    fields: [image, title, subtitle, description]
    image-height: "150px"
    categories: false
    feed: false
    filter-ui: false
    sort-ui: false
---

```{r setup, include=FALSE}
source("R/helpers.R")
```

My teaching interests complement my research and span the theoretical, empirical, and policy-oriented foundations of economics: microeconomic and macroeconomic theory, econometrics, mathematical economics and optimization, and monetary economics.

I teach across the undergraduate, postgraduate, and doctoral programmes, at the Department of Economics, Aligarh Muslim University. 

Alongside the theory courses, I teach computational and econometric methods, including time series, panel data, financial econometrics, and research methodology. I try to integrate teaching these courses in R, with an emphasis on reproducibility. 


## Course Websites

Slides, datasets and worked examples for the courses that have their own
material online. Everything is free to use.

::: {#course-pages}
:::

```{r teaching, results="asis"}
print_teaching()
```

*Earlier (2014–2018): Assistant Professor, Department of Economics, MANUU, Hyderabad.*

```{r advising, results="asis"}
print_advising()
```
