---
title: "Title of Document"
author:
  - "**Student Name**"
  - 24ECM05
date: "`r Sys.Date()`"
format: 
  docx:
    toc: true
    toc-depth: 3
    toc-location: body
    toc-title: Table of Contents
    number-sections: true
    highlight-style: github
    theme: dark
    fontsize: 12pt
editor: visual
---

```{css, echo = FALSE}
.justify {
  text-align: justify !important
}
```

{{< pagebreak >}}

## Introduction

Let me first talk about the structure of the project, then I will talk about writing about it in the **Quarto**. The project should include following sections.

-   Introduction
-   Literature Review
-   Data and Methodology
    -   Data
    -   Methodology
-   Results Analysis
-   Conclusion
-   References

I am sharing sample papers to understand, what should go in these sections.

## Writing in Quarto

Now let me introduce different procedures needed to complete the project in **Qaurto**.

-   First **YAML**

In **YAML**, we include about the document information and formatting structure of the document. Here I have included bare minimum **YAML**, which will be sufficient to complete the project in a structured format.

-   Inserting a link

::: justify
To include a link, you have two options. One where you want to display the full link then write like this <https://www.google.co.in/>. Remember the link should be in between the `<>` if you want to show a text instead of a full link and hide the link behind the text. Then write like this. To go to Google, click on the word [Google](https://www.google.co.in/).
:::

-   Different formats of texts

To write a **bold text**, you have to the texts in between the symbols `** **,` and to write ***italic texts***, you have to write the texts in between the symbols `*** ***`.

-   Bullet items and numbered items.

First, to create bullet items you can use the symbol `*` or `-`. You must give space after the symbol and also have space between one bullet item to another. To create a sub-bullet, you have to shift the next bullet item towards the right. For example, see this list.

-   item1

-   item2

    -   sub-item1

    -   sub-item2

-   Numbered items

1.  item1

2.  item2

    i)  sub-item1

    ii) sub-item2

You may use the tab to create a sub-bullet(Use two tab spaces to create sub-numbered items).

## Writing equations

There are two types of equations you may need to insert in the document. One is known as an **inline equation,** such as $y = a + b$. The second is known as **Display equations,** such as

$$
y = \alpha + \beta x
$$

::: callout-warning
## Note

To write a display equation, use `$$` sign at the beginning of the equation and at the end of the equation, and there should not be any **space** after the `$$` sign and before the `$$` sign.
:::

To write a numbered equation, use a tag after the `$$` sign at the end of the equation like this.

$$
y = \alpha + \beta x
$$ {#eq-one}

::: callout-warning
## Note

The tag must be inside the curly bracket `{}` and start with `#eq,` and after `-` you can put any identification. For example, I have used the word **one**.
:::

In quarto, we use `LaTeX` language to write `mathematical symbols and equations`. To see the full list of `LaTeX` code, click [HERE](https://www.cmor-faculty.rice.edu/~heinken/latex/symbols.pdf).

## Changing the colour of the text.

You change the colour of the text by using the command `{style="color: red"}`. To make the colour of the text [blue]{style="color: blue"}, I have used `{style="color: blue"}`. Similarly, to use the color of the text [red]{style="color: red"}, I have used `{style="color: red"}`. The specific text of which you want to change the colour must be within `[]` and followed by the style.

## Inserting code

To insert the code chunk (group of codes), you have to use the **back quote (\`)**, just below the **twiddle/tilde (\~)** symbol. To insert a code chunk, in the beginning, put three \***back quote(\`\`\`)** symbols and at the end also put three **back quote(\`\`\`)** symbols.

### Some examples

Here we will import the data for further analysis.

::: callout-warning
## Note

Another and easiest way to insert code chunk is to switch to the visual editor and from the insert tab click on the **executable cell** or use shortcut keys `Ctrl+Alt+I`.
:::

```{r}
#| echo: false
#| eval: true
data <- read.csv("Data.csv")

```

In the above code chunk use `echo: false` will make sure that code is not printed in the document. However, `eval: true` makes this code executable that is data will be imported and stored in dataframe named `data`.

Lets examine the data by using code `head()`.

```{r}
#| echo: false
#| eval: true
head(data,3)
```

The results will be printed in the document.

Now lets make line graph of these variables.

```{r}
#| echo: false
#| eval: true
#| fig-width: 6
#| fig-height: 5
#| fig-align: center
#| fig-cap: "A Figure"
#| fig-cap-location: bottom
#| label: fig-line-plot
plot(data$Year, data$SID,
     type = "l",
     lwd = 2, col = "red",
     main = "Line graph",
     ylab = "Values of HI",
     xlab = "Year")
```

Here in the above code chunk `fig-width` and `fig-height` is for the width and height of the figure respectively. `fig-align` is for the alignment of figure, which is in center of the document width. Similarly `fig-caption` and `fig-cap-location` is respectively for the caption of the figure and position of caption. Here option `label` makes the figure cross-referenceable. Here we cross-reference the above figure like this, the @fig-line-plot shows overtime growth of HI.

::: callout-warning
## Note

First to `lable` the figure command must start with `fig` and then after `-` name the graph/plot. For example here name is `line-plot`. Then use \@ followed by `fig-line-plot` to cross-reference the figure.
:::

```{r}
#| label: tbl-des
#| echo: false
#| tbl-cap: "Descriptive Statistics"

library(knitr)
library(psych)
subdata <- data[,c("HI","SID")]
kable(describe(subdata, IQR=FALSE, ranges = FALSE))
```

Here `kable` is to make the table in the document. You can cross-reference the table as @tbl-des.

## Final words

This is a sample, you can explore further for the other options. You may find the complete list of option [HERE](https://quarto.org/docs/reference/formats/docx.html).

::: callout-warning
## Note

Remaining formatting, you may do in the generated word file also.
:::
