---
title: "Tidyverse"
author: "Md. Zulquar Nain"
format: 
  revealjs:
    scrollable: true
    incremental: true
    slide-number: c/t
    theme: dark
    preview-links: auto
    chalkboard: 
      boardmarker-width: 5
editor: visual
---

## What is Tidyverse?

-   **The Tidyverse** is a collection of R packages for data science that share a common design philosophy, grammar, and data structures.

-   It makes it easy to install and load core packages with a single command.

-   It includes packages that are used in everyday data analysis.

## Why use Tidyverse ?

-   **Consistency**: Consistent syntax and data structures across packages.

-   **Ease of Use**: Simple, intuitive functions.

-   **Performance**: Efficient and optimized for modern data science.

-   **Versatility**: Ideal for various tasks in data manipulation, visualization.

-   **Comprehensive**: Complete package for all stages of data science workflows.

## Key Packages in Tidyverse {.scrollable}

-   **`ggplot2`**: Data visualization.
-   **`dplyr`**: Data manipulation.
-   **`tidyr`**: Data tidying.
-   **`readr`**: Data input.
-   **`purrr`**: Functional programming.
-   **`tibble`**: Data frames.
-   **`stringr`**: String manipulation.
-   **`forcats`**: Factor handling.
-   **`lubridate`**: Date and time manipulation.

# Installation and Setup

-   To install Tidyverse: install.packages("tidyverse")

-   To load Tidyverse into your current R session: library(tidyverse)

## Pipe Operator( %\>% ) {.scrollable}

-   The pipe operator **(%\>%)** or **(\|\>)** is used to pass the result of one function directly to the next function in a sequence, making your code more readable and intuitive.

-   Pipes allow you to write operations sequentially, enhancing clarity and reducing complexity.

-   Syntax:

    `data %\>% function1() %\>% function2() %\>% function3()`

-   **data**: The input data is passed through the pipe.

-   **function1(), function2(), function3()**: These functions operate on the data in sequence, with the output of one function feeding directly into the next.

-   The shortcut to type the pipe operator in RStudio is given by CTRL/CMD + Shift + M.

------------------------------------------------------------------------

## Example of pipe operators {.scrollable}

Simulating a sample of data by using the function sample we draw randomly (without replacement) 5 numbers between 1 and 20 (1:20) and compute the log transformation of the vector.

. . .

```{r}
#| echo: true
#| output-location: fragment

# without pipe operator
set.seed(44)
x = sample(1:20, 5)
x

```

. . .

```{r}
#| echo: true
#| output-location: fragment
log(x)
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#with pipe operator
x |> log()
#it's also possible to omit the parentheses given that there is no input


```

------------------------------------------------------------------------

# Package `dplyr`

## Data Manipulation with `dplyr` {.scrollable}

-   dplyr is designed for easy data manipulation using verbs that describe the operations you want to perform. These common functions are:

-   `filter()` : Select rows based on conditions.

-   `select()` : Choose columns to keep.

-   `mutate()`: Add new columns or modify existing ones.

-   `arrange()`: Sort the data.

-   `summarize()`: reduces multiple values down to a single summary

-   `group_by()`: Aggregate data by groups.

------------------------------------------------------------------------

## `dplyr` verbs {.scrollable}

Taking *diamonds* dataset of the prices and other attributes of almost 54,000 diamonds **(see ?diamonds)**.

. . .

```{r}
#| echo: true
#| output-location: fragment

# understand the dataset of diamonds
library(ggplot2)
class(diamonds)

```

. . .

```{r}
#| echo: true
#| output-location: fragment

str(diamonds)
```

. . .

```{r}
#| echo: true
#| output-location: fragment

# we can also use tidyverse function
library(dplyr)
glimpse(diamonds)

```

------------------------------------------------------------------------

## `select()` {.scrollable}

. . .

```{r}
#| echo: true
#| output-location: fragment
diamonds %>% select(carat)
```

. . .

-   Selecting more than one column :

. . .

```{r}
#| echo: true
#| output-location: fragment
diamonds %>% select(carat, cut, color, price)
```

-   Alternate choices

. . .

```{r}
#| echo: true
#| output-location: fragment


diamonds %>% select(carat : color, price)
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#selcting columns starting with 'c'
diamonds %>% select(starts_with("c"))
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#select all the columns but carat
diamonds %>% select(-carat)
```

-   Now you do this select all the columns but not the ones with a name starting with “c”

. . .

```{r}
#| echo: true
#| output-location: fragment
diamonds %>% select(- starts_with("c"))
```

------------------------------------------------------------------------

## `filter` {.scrollable}

. . .

```{r}
#| echo: true
#| output-location: fragment

diamonds %>% filter(cut == "Premium")
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#include more conditions by using '&' or 'comma'
diamonds %>% filter(cut == "Premium" & color == "D") 
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#The output of a selection can be saved in a new data frame
myselection = diamonds %>% filter(between(price, 500,600))
```

------------------------------------------------------------------------

## `summarise` {.scrollable}

. . .

```{r}
#| echo: true
#| output-location: fragment

#mean and median of price.

diamonds %>% 
  summarise(mean(price),
            median(price))
```

. . .

```{r}
#| echo: true
#| output-location: fragment

#the number of diamonds with a price > 15000$
diamonds %>% 
  summarise(veryexp = sum(price > 15000),
            veryexpprop = mean(price>15000),
            veryexpperc = mean(price>15000)*100)
```

------------------------------------------------------------------------

## `group_by` {.scrollable}

. . .

-   **`group_by()`** is used to group rows of a data frame by one or more columns.

-   This helps in performing operations like summarizing or aggregating data by categories.

. . .

```{r}
#| echo: true
#| output-location: fragment

diamonds %>% 
  group_by(cut,color) %>% 
  summarise(mean(price)) 

```

------------------------------------------------------------------------

## `mutate` {.scrollable}

-   used to create new column or modify existing columns in the data frame

. . .

```{r}
#| echo: true
#| output-location: fragment

newdiamonds = diamonds %>% 
  mutate(newcol = ifelse(price < 1000, "Yes", "No"))
# Create a new column to categorize products as "Yes" if the price is less than $1000, and "No" otherwise.
```

. . .

```{r}
#| echo: true
#| output-location: fragment

# Derive the frequency distribution of 'newcol' along with percentages.
newdiamonds %>% 
  count(newcol) %>% 
  #summarise(perc=n/nrow(newdiamonds)*100)
  mutate(perc=n/nrow(newdiamonds)*100)
```

------------------------------------------------------------------------

## `arrange` {.scrollable}

. . .

```{r}
#| echo: true
#| output-location: fragment

#sort diamonds according to price
diamonds %>% 
  arrange(price) %>% 
  tail
```

. . .

```{r}
#| echo: true
#| output-location: fragment

diamonds %>% 
  arrange(desc(price))
```

------------------------------------------------------------------------

## Example with dplyr verbs combined {.scrollable}

-   Taking 'mtcars' and filtering rows where mpg \> 20, select specific columns, create a new column, and arrange the data by mpg in descending order.

. . .

```{r}
#| echo: true
#| output-location: fragment

library(dplyr)

mtcars %>% 
  filter(mpg > 20) %>%
  select(mpg, wt, hp) %>% 
  mutate(mpg_per_weight = mpg / wt) %>% 
  arrange(desc(mpg))

```

------------------------------------------------------------------------

# Thanks
