---
title: "Tidyverse Other packages"
author: "Dr. Md. Zulquar Nain"
format: 
  revealjs:
    incremental: true
    slide-number: c/t
    theme: dark
    preview-links: auto
    chalkboard: 
      boardmarker-width: 5
editor: visual
---

## tidyr - Data Tidying {.scrollable}

::: incremental
-   A package in the Tidyverse for tidying data

-   Helps convert data between wide and long formats

-   Simplifies reshaping and cleaning datasets

-   Key Functions in tidyr
:::

. . .

### pivot_longer() {.incremental}

Converting from Wide data *(multiple columns for the same variable)* to Long data *(one column for each variable)*

```{r}
#| echo: true
#| output-location: fragment

library(tidyr)
  
  data_wide <- tibble(
    id = 1:3,
    `2021` = c(100, 150, 200),
    `2022` = c(110, 160, 210)
  )
  
  data_long <- data_wide %>%
    pivot_longer(cols = `2021`:`2022`, names_to = "year", values_to = "value")
  
  print(data_long)
```

### pivot_wider()

Converting from Long to Wide Format

```{r}
#| echo: true
#| output-location: fragment

data_long <- tibble(
  id = c(1, 1, 2, 2, 3, 3),
  year = c("2021", "2022", "2021", "2022", "2021", "2022"),
  value = c(100, 110, 150, 160, 200, 210)
)

data_wide <- data_long %>%
  pivot_wider(names_from = year, values_from = value)

print(data_wide)

```

### separate()

Splits a single column into multiple columns

```{r}
#| echo: true
#| output-location: fragment

data <- tibble(name = c("John Doe", "Jane Smith", "Alice Johnson"))

data_separated <- data %>%
  separate(name, into = c("first_name", "last_name"), sep = " ")

print(data_separated)
```

### unite()

Combines multiple columns into one

```{r}
#| echo: true
#| output-location: fragment

data <- tibble(first_name = c("John", "Jane", "Alice"),
               last_name = c("Doe", "Smith", "Johnson"))

data_united <- data %>%
  unite("full_name", first_name, last_name, sep = " ")

print(data_united)
```

### drop_na()

Removes rows with missing values

```{r}
#| echo: true
#| output-location: fragment

data <- tibble(x = c(1, NA, 3), y = c(4, 5, NA))

data_clean <- data %>%
  drop_na()

print(data_clean)
```

### fill()

Fills missing values using the previous or next available value

```{r}
#| echo: true
#| output-location: fragment

data <- tibble(x = c(1, NA, 3, NA, 5))

data_filled <- data %>%
  fill(x, .direction = "down")

print(data_filled)
```

------------------------------------------------------------------------

## tidyr vs dplyr

-   **tidyr**: Specializes in reshaping and tidying data (e.g., pivot_longer(), pivot_wider(), separate()).

-   **dplyr**: Specializes in data manipulation, such as subsetting, grouping, and summarizing (e.g., filter(), mutate(), summarize()).

------------------------------------------------------------------------

## readr - Data Import/Export {.scrollable}

-   part of the tidyverse, focused on reading and writing rectangular data (e.g., CSV, TSV).

-   Faster and more consistent than base R functions

-   Makes data handling easy and efficient.

-   Key Functions:

. . .

### read_csv()

```{r}
#| echo: true
#| output-location: fragment

library(readr)
data_csv <- read_csv("hsbraw.csv")
head(data_csv)
```

### read_delim()

```{r}
#| echo: true
#| output-location: fragment

# Read a pipe-separated file
data_pipe <- read_delim("hsbraw.txt", delim = "|")
head(data_pipe)
```

### write_csv()

```{r}
#| echo: true
#| output-location: fragment

# Write a data frame to a CSV file
write_csv(data, "hsbraw.csv")
```

### write_delim()

```{r}
#| echo: true
#| output-location: fragment

# Write data to a pipe-separated file
write_delim(data, "hsbraw.txt", delim = "|")
```

------------------------------------------------------------------------

## purrr - Functional Programming {.scrollable}

-   enhances R's functional programming capabilities by providing a consistent, simple way to apply functions to lists, vectors, and other data structures.

-   Key function:-

. . .

### map()

Applies a function to each element of a list and returns a list.

```{r}
#| echo: true
#| output-location: fragment

library(purrr)

# Create a list of numbers
numbers <- list(1, 2, 3, 4, 5)

# Apply a function to square each number using map
squared_numbers <- map(numbers, ~ .x^2)

# Print the result
squared_numbers

```

### map_dbl()

To apply a function and return a vector (not a list), map_dbl() is used.

```{r}
#| echo: true
#| output-location: fragment

# Apply the same function, but return a vector of doubles
squared_numbers_vector <- map_dbl(numbers, ~ .x^2)

# Print the result
squared_numbers_vector
```

### map2()

-   Applies a function to two inputs simultaneously.

-   useful when you need to operate on two lists.

```{r}
#| echo: true
#| output-location: fragment

# Create two lists
list1 <- list(1, 2, 3)
list2 <- list(10, 20, 30)

# Use map2 to add corresponding elements
sum_list <- map2(list1, list2, ~ .x + .y)

# Print the result
sum_list
```

### map_chr()

Used to get a character vector instead of a numeric one.

```{r}
#| echo: true
#| output-location: fragment

# Create a list of numbers
numbers2 <- list(1, 2, 3, 4)

# Use map_chr to convert each number to a character string
char_numbers <- map_chr(numbers2, ~ as.character(.x))

# Print the result
char_numbers
```

------------------------------------------------------------------------

## purrr vs tidyr

-   **purrr** is specifically focused on working with lists and vectors using functional programming techniques.

-   **tidyr** is focused on reshaping data frames (long to wide format and vice versa), separating and combining columns.

------------------------------------------------------------------------

## tibble - Modern Data Frames {.scrollable}

-   A modern version of a data frame.

-   More robust and user-friendly than traditional data frames.

-   Improved handling of large data sets

. . .

### Basic Tibble

```{r}
#| echo: true
#| output-location: fragment

# Create a basic tibble
people_tibble <- tibble(
  Name = c("John", "Alice", "Bob"),
  Age = c(25, 30, 22),
  Location = c("New York", "London", "Paris")
)

# Print the tibble
people_tibble
```

### Tibble with Mixed Data Types

```{r}
#| echo: true
#| output-location: fragment

# Create a tibble with mixed data types
mixed_tibble <- tibble(
  Name = c("John", "Alice", "Bob"),
  Age = c(25, 30, 22),
  Is_Active = c(TRUE, FALSE, TRUE),
  Height = c(5.9, 5.6, 6.1)
)

# Print the tibble
mixed_tibble
```

------------------------------------------------------------------------

### Tibble vs. Data Frame

Data Frame

::: incremental
-   Displays all rows, which can be overwhelming with large datasets.
-   Automatically converts character vectors to factors (unless specified otherwise).

Tibble

-   Only shows the first few rows, making it easier to handle large datasets.

-   Does not convert character columns into factors by default, avoiding a common issue with data frames.
:::

------------------------------------------------------------------------

## stringr - String manipulation {.scrollable}

-   Provides simple functions for common text operations.

-   Focuses on consistency and ease of use.

-   Key Functions:-

. . .

### str_detect()

```{r}
#| echo: true
#| output-location: fragment

# checks if a pattern exists in the string
library(stringr)
text <- "Hello, world!"
has_hello <- str_detect(text, "Hello")
has_hello
```

### str_replace()

```{r}
#| echo: true
#| output-location: fragment

#replaces the first occurrence of a pattern with a new string
text <- "I love R programming."
new_text <- str_replace(text, "R", "Python")
new_text
```

### str_replace_all()

```{r}
#| echo: true
#| output-location: fragment

# replaces all occurrences of a pattern
text <- "The cat is on the mat. The cat is cute."
new_text_all <- str_replace_all(text, "cat", "dog")
new_text_all
```

### str_sub()

```{r}
#| echo: true
#| output-location: fragment

#extracts a substring from a string
text <- "Hello, world!"
substring <- str_sub(text, 1, 5)
substring
```

### str_to_lower()

```{r}
#| echo: true
#| output-location: fragment

# converts all characters in the string to lowercase
text <- "HELLO, WORLD!"
lowercase_text <- str_to_lower(text)
lowercase_text
```

### str_split()

```{r}
#| echo: true
#| output-location: fragment

#splits a string into a list by a specified delimiter 
text <- "apple,banana,cherry"
split_text <- str_split(text, ",")
split_text
```

### str_length()

```{r}
#| echo: true
#| output-location: fragment

#returns the number of characters in the string
text <- "Hello"
string_length <- str_length(text)
string_length
```

------------------------------------------------------------------------

## lubridate {.scrollable}

-   It makes working with date-times easier

-   Provides functions to manipulate, parse, and format date-time data.

-   Simplifies operations like extracting parts of date-time, arithmetic operations, and handling time zones.

. . .

### Parsing Dates

```{r}
#| echo: true
#| output-location: fragment

library(lubridate)
# Parse a date in year-month-day format
date1 <- ymd("2025-03-03")
print(date1)
```

### Parsing Dates with Time

```{r}
#| echo: true
#| output-location: fragment

# Parse a date-time with time and time zone
datetime1 <- ymd_hms("2025-03-03 12:30:45")
print(datetime1)
```

### Extracting Date-Time Components

```{r}
#| echo: true
#| output-location: fragment

# Extract components from a date-time object
year(datetime1)
month(datetime1)
day(datetime1)
hour(datetime1)
minute(datetime1)
second(datetime1)
```

### Handling Time Intervals

```{r}
#| echo: true
#| output-location: fragment

# Define an interval
start <- ymd_hms("2025-03-01 00:00:00")
end <- ymd_hms("2025-03-05 23:59:59")
interval1 <- interval(start, end)
print(interval1)
```

------------------------------------------------------------------------

## forcats - Working with Factors {.scrollable}

-   Makes working with categorical variables easier.

-   Provides tools to manipulate factors and handle tasks like reordering, renaming, and combining levels.

-   Key functions :

. . .

### Creating Factors

```{r}
#| echo: true
#| output-location: fragment

library(forcats)
# Create a factor from a vector
categories <- c("Low", "High", "Medium", "Low", "High")
factor_categories <- factor(categories)
print(factor_categories)
```

### Reordering factor levels

```{r}
#| echo: true
#| output-location: fragment

# Reorder factor levels
factor_ordered <- fct_reorder(factor_categories, c(3, 2, 1, 4, 5))
print(factor_ordered)
```

### Changing Factor Levels

```{r}
#| echo: true
#| output-location: fragment

# Change factor levels
new_factor <- fct_recode(factor_categories, "Very Low" = "Low", "Very High" = "High")
print(new_factor)
```

### Combining Factor Levels

```{r}
#| echo: true
#| output-location: fragment

# Combine similar factor levels
collapsed_factor <- fct_collapse(factor_categories, 
                                 "Low/Medium" = c("Low", "Medium"),
                                 "High" = "High")
print(collapsed_factor)
```

### Working with Factor Levels

```{r}
#| echo: true
#| output-location: fragment

# Drop unused factor levels
dropped_factor <- fct_drop(factor_categories)
print(dropped_factor)
```

### Visualizing Factor Data

```{r}
#| echo: true
#| output-location: fragment


library(ggplot2)
# Create a plot with reordered factor levels
ggplot(mpg, aes(x = fct_rev(class))) + 
  geom_bar() + 
  labs(title = "Count of Cars by Class")
```

------------------------------------------------------------------------

## forcats vs dplyr

-   **forcats** works with factors used for modifying factor levels, whereas **dplyr** also provides some functions that manipulate factors (e.g., mutate() and factor()).

-   **forcats**: Designed specifically for working with factors and making factor level manipulations easier.

-   **dplyr**: While dplyr can work with factors, forcats is more specialized for factor-specific operations.

------------------------------------------------------------------------

## THANKS {.center}
