---
title: Data Types & Structures in `R`
author: Md Zulquar Nain
institute: AMU
date: today
date-format: long
format: 
  revealjs:
    theme: night
   # logo: imgages/logo.png 
    incremental: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    preview-links: auto
  pdf: default
editor: visual
from: markdown+emoji
---

# Data Types

## Objects {.scrollable}

-   Most statistical software (e.g., SPSS, Stata) operates on datasets, which consist of rows of observations and columns of variables
-   R is an `object-oriented` programming language (like Python,JavaScript).
-   So, what is an `object`?
-   In computer language - complex definition
-   Anything to which you assigned a value

```{r}
a <- 5
d <- "You are SMART!"
```

. . .

::: {.callout-important style="color: red"}
**Objects are like boxes in which we can put things: data, functions, and even other objects.** Ben Skinner
:::

## Data Objects {background="#4682B4"}

:::: r-fit-text
::: centre
-   Data usually obtained from external sources
-   Collected by researchers using survey
-   In `R` import those data files into `R Objects`
:::
::::

## Data Types {.scrollable}

::: centre
-   Use function `typeof()` to check the type of data
-   the function `typeof()` returns the **Data type** or **data structure**
-   Data Types are
    -   **Double**
    -   **Integer**
    -   **Complex**
    -   **Logical**
    -   **Character**
    -   **Factor**
    -   **Date and time**
:::

## Data Types

::: centre
-   **Double**
    -   `Doubles` are numbers like 2.0, 2.2, 2.999
    -   May or may not include decimals
    -   Mostly used to represent continuous variables like wage, height, age
    -   In `R` by default all numbers are `Double`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
x <- 10.5
is.double(x)

```
:::

## Data Types

::: centre
-   **Integer**
    -   `Integers` are natural numbers
    -   Mostly used for counting variables
    -   However, in `R` by default treated as `Double`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
x <- 6
typeof(x)

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
x <- as.integer(9)
typeof(x)

```
:::

## Data Types

::: centre
-   **Complex**
    -   `Complex` like $x+yi$
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
x <- -1
typeof(x)
sqrt(x)

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment

# Complex
y <--1+(0+0i)
sqrt(y)

```
:::

## Data Types

::: centre
-   **Logical**
    -   `Logical`: a variable of logical type data has values like `TRUE` or `FALSE`
    -   Usually used to represent conditional statements
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
x <- 20
y <- 15
a <- x > y
a
typeof(a)
```
:::

## Data Types

::: centre
-   **Character**
    -   `Characters` represents a `string values` in `R`
    -   Usually a string or collection of characters are kept in`double quotes`
    -   Anything in a double quote is considered a string in `R`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
x <- "You are smart"
print(x)
y <- "5"
typeof(y)
```
:::

## Data Types {.scrollable}

::: centre
-   **Factor**
    -   `Factor` represents categorical data
    -   For example gender(Male, Female), employed and unemployed
    -   Also handy in case of panel and longitudinal data
    -   Factor objects can be created from character or numeric objects
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
# Created from Characters
cat <- c("A","B","C")
# Use functions to convert into factor type
cat=as.factor(cat)
cat 
levels(cat)
```
:::

## Data Types {.scrollable}

::: centre
-   **Date and Time**
    -   We need quite often especially when dealing with time series models
    -   Function `as.Date` is used to create `Date and Time` objects\
    -   See `help(as.Date)`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
today <-"25-03-2023" 
today <- as.Date(today,"%d-%m-%Y")
today
today1 <- Sys.Date()
today1
today2 <- format(today1, "%d %B %Y")
today2

```
:::

## Missing Values in `R`

::: centre
-   In `R` missing data is represented by `NA` meaning **Not Available**
-   Another symbol appears in `R` is `NaN` meaning **Not a Number**
-   `NULL` in `R` represent an object with ZERO length
-   `-Inf` and `Inf` represents negative and positive infinity
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| code-line-numbers: true
#| output-location: fragment
# Create a vector using function `c()`
ownd <- c("10","30","missing")
# converting to double
ownd <- as.double(ownd)
  # check for the missing value
is.na(ownd)

```
:::

# Data Structure in `R`

## Data Structure in R {.scrollable}

-   For analysis,data should be structured in well-defined manner.
-   In `R` there are certain structures followed by imported data.
    -   Vector
    -   Matrices
    -   Lists
    -   Array
    -   Data Frame

## Vectors

:::: r-fit-text
::: centre
-   scalars Vs Vectors?
-   groups of values of same data types
-   Simplest function to define a vector `c(value1,value2,...)`
-   All the operators and functions discussed earlier can be used for `vectors` also
-   However, operations are performed element by element
:::
::::

. . .

```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# define a vector `vec` with function `c`
vec <- c(2, 4,-4,0)
vec
print(vec)

```

## Vectors {background="#4682B4"}

::: centre
-   Other way to generate vectors
:::

. . .

| Functions           | Explanations                          |
|---------------------|---------------------------------------|
| `numeric(n)`        | vector with `n zeros`                 |
| `rep(x,n)`          | Vector with `n` equal elements of `x` |
| `seq(x)/seq(1:x)`   | Sequence from `1` to `x`              |
| `seq(f,x)/seq(f:x)` | Sequence from `1` to `x`              |
| `seq(f,x,s)`        | Sequence from `1` to `x` in steps `s` |

: Special functions to create vectors

## Basic Operations of a Vector {.scrollable}

|  Functions  | Explanations                          |
|:-----------:|---------------------------------------|
| `length(v)` | Number of elements in vector `v`      |
|  `max(v)`   | Largest Number in vector `v`          |
|  `min(v)`   | Smallest number in vector `v`         |
|  `sum(v)`   | Sum of the elements in vector `v`     |
|  `prod(v)`  | Product of elements in vector `v`     |
|  `sort(v)`  | Sorting of the elements of vector `v` |

## Basic Operations of a Vector {.scrollable}

```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# define a vector `vec` with function `c`
vec <- c(2, 4,-4,0,1,3)
vec
#length
length(vec)
# largest number
max(vec)
#smallest number
min(vec)
#sum of elements
sum(vec)
#product of elements
prod(vec)
# sorting
sort(vec)

```

## Special Types of Vectors {.scrollable}

::: centre
-   Character Vectors
-   Logical Vectors
-   Factors
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Character Vectors
sname <- c("Raju","abdul","Soha")
sname

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# logical Vectors
a <- c(7, 2, 6, 10, 4, 1, 3)
a
b <- a<3 | a>=6
b

```
:::

## Special Types of Vectors {.scrollable}

::: centre
-   Factors
-   Rate my teaching:joy:
-   Good=1, Very good=2, Best=3
-   ten students' response are recorded in numbers
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# ratings in numbers
qt <- c(3, 2, 1, 1, 2, 3, 2, 1, 3, 3)
labelnames <- c("good", "very good", "best")
#creating a factor vector
fqt <- factor(qt,labels = labelnames)
qt
fqt

```
:::

## Matrices {.scrollable}

::: centre
-   A matrix is a collection of data elements arranged in a two-dimensional rectangular layout.
-   all the elements in a matrix are of the same data type
-   Function `matrix` is used to create matrix
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

m1 <- matrix(c(1, 2, 3, 4, 5, 6), nrow= 3, ncol = 2, byrow = TRUE)
# nrow- specify number of rows
# ncol-specify the number of columns
#byrow- fill the matrix in rows with the data supplied 
m1


```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment


m2 <- matrix(c(1, 2, 3, 4, 5, 6), nrow= 3, ncol = 2, byrow = FALSE)
m2

```
:::

## Other functions to create matrices {.scrollable}

-   Other ways to create matrix
-   By using functions `dim`, `cbind` and `rbind`

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Use of `dim` function
x <- 1:12
dim(x) <- c(3,4) 
x

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Use of `dim` function
dim(x) <- c(4,3)
x

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Use of `rbind` function
# rows must have same length 
r1 <- 1:3
r2 <- c(4, 5, 6)
rm <- rbind(r1, r2)  
rm 

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Use of `cbind` function
# columns must have same length 
c1 <- 1:3
c2 <- c(10, 9, 8)
c3 <- c(20, 5, 15)
cm <- cbind(c1, c2, c3)  
cm 

```
:::

## Names and indexing in Matrices {.scrollable}

-   Column names and row names
-   Extracting a specific element(s)

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
cm 

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Names of columns and rows
colnames(cm) <- c("Eco", "Hist", "Pol")

rownames(cm) <- c("SI", "SII", "SIII")
cm


```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Extracting `Pol` in `SIII`
cm[3,3]


```
:::

. . .

-   Extract first row and all columns
-   Extract second column and all rows
-   Extract all rows, second and third columns

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Extract first row and all columns
cm[1,]

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Extract second column and all rows
cm[,2]

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# Extract first row and second column
cm[,c(2,3)]

```
:::

## Matrix manipulations {.scrollable}

::: centre
-   all the mathematical functions available for vectors are applicable on a matrix
-   all operations are applied on each element in a matrix
-   Multiply the matrix `m1` with `3`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# ratings in numbers
m1 <- matrix(c(1, 2, 3, 4, 5, 6), nrow= 3, ncol = 2, byrow = TRUE)
# Multiplying matrix m1 by 3
m3 <- m1*3
m1
m3
```
:::

. . .

-   Mathematical operations-for example matrix multiplication
-   Rule of dimensionality must satisfy

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
# ratings in numbers
m1 <- matrix(c(1, 2, 3, 4, 5, 6), nrow= 3, ncol = 2, byrow = TRUE)
# Multiplying matrix m1 by 3
m1
dim(m1)

```
:::

. . .

-   create another matrix with `2 rows` and `3 columns`

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

m4 <- matrix(c(1, 2, 3, 4, 5, 6), nrow= 2, ncol = 3, byrow = TRUE)
dim(m4)
# Multiplying matrix m1 with m4 
m5 <- m1%*%m4
m5

```
:::

## Lists {.scrollable}

::: centre
-   In `R` a **list** is a generic collection of objects
-   Each component can have different types of data
-   `mylist <- list(name1=component1, name2=component2, ...)`
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# generate a list of object

mylist <- list(A=seq(10, 30,5), student="aadil", idm=diag(3))

#print my list

mylist

```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# generate a list of object

mylist <- list(A=seq(10, 30,5), student="aadil", idm=diag(3))

#print vector names
names(mylist)


```
:::

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# generate a list of object

mylist <- list(A=seq(10, 30,5), student="aadil", idm=diag(3))

#print component A
mylist$A


```
:::

## Data Frames {.scrollable}

-   An object consisting of several variables
-   Rectangular shape
-   rows representing observational units
-   columns representing the variables
-   Is it matrix?
-   No
-   Matrix include only one type of data
-   Data frame can include different types of data
-   Use the command `data.frame` or `as.data.frame` to transform into data frame

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Creating a data frame

# Define a vector 
year <- c( 1991, 1992, 1993, 1994, 1995)

GDP <- c(200, 250, 300, 320, 400)

INV <- c(100, 150, 100, 120, 200)

macro_mat <- cbind(GDP, INV)

rownames(macro_mat) <- year

macro_mat

# create data frame

macro <- as.data.frame(macro_mat)
macro
```
:::

## Accessing Data frame {.scrollable}

-   Accessing a single variable
-   Generating a new variable

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment

# Existing data frame
macro
#Accessing the GDP data Only
macro$GDP

# Generating a new variable in the data frame
macro$lnGDP <- log(macro$GDP)

# Using `with` function
macro$lnINV <- with(macro,log(INV))

#Using `attach()` function
attach(macro)
macro$total <- GDP+INV
detach(macro)

# Results
macro


```
:::

## Creating a subset of data frame {.scrollable}

::: fragment
```{r,message=FALSE}
#| echo: true
#| eval: TRUE
#| code-overflow: scroll
#| code-fold: false
#| code-summary: "The Code"
#| output-location: fragment
#Existing data frame
macro

# Taking subset
subdata <- subset(macro, total>400)

subdata
```
:::

#  {background-image="images/foggy.webp"}

:::: r-fit-text
::: center
[**THANKS**]{style="color: red;"}
:::
::::
