---
title: "Descriptive Statistics"
author: "Md Zulquar Nain"
format: 
  revealjs:
    incremental: true
    scrollable: true
    slide-number: c/t
    theme: dark
    preview-links: auto
    chalkboard: 
      boardmarker-width: 5
editor: visual
---


## Importing the Data File

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| evaluation: false
# importing data from `csv` file
datai <- read.csv("hsbraw.csv")
```

-   `datai` - name of the imported data file in`R`

-   `hsbraw.csv` name of the `csv` file being imported

## Exploring the Dataset I {background="#4682B4"}

-   Class, structure and dimension of the dataset

. . .

```{r }
#| echo: true
#| output-location: fragment

# Structure of the data
str(datai)
```

. . .

```{r }
#| echo: true
#| output-location: column-fragment


#Class of the data
class(datai)

```

. . .

```{r }
#| echo: true
#| output-location: column-fragment

# Dimension of the data
dim(datai)


```

## Exploring the Dataset II {background="#4682B4"}

-   First `n` rows of observations of the data set
    -   `head(data.frame name, n)`
-   Last `n` rows of observations of the data set
    -   `tail(data.frame name, n)`

. . .

```{r}
#| echo: true
#| output-location: fragment

# View top two rows of the data
head(datai,2)
```

. . .

```{r}
#| echo: true
#| output-location: fragment

# View bottom two rows 
tail(datai,2)
```


# Descriptive Statistics {background="#13464F"}

## Measures of Central tendency {background="#13464F"}

-   For continuous variables are the
    -   mean, median, and variance
-   Functions in `R`
    -   `mean()`
    -   `median()`
    -   `var()`
    -   `sd()` for standard deviation

## Descriptive statistics

-   `summary()` function available with base `R`
-   `mean`,`median`,`25th` and `75th` quartiles
-   `min` and `max`

. . .

```{r}
#| echo: true
#| output-location: fragment
summary(datai)
```

## Descriptive statistics

-   Selecting a specific column

. . .

```{r}
#| echo: true
#| output-location: fragment
summary(datai$read)

```

-   More than one column

. . .

```{r}
#| echo: true
#| output-location: fragment
# create a subdata 
subdata <- datai[,c("read","write","math")]
# summary of the subset
summary(subdata)

```

## Descriptive Statistics

-   Using `describe` function form the `psych` package

-   More control

. . .

```{r}
#| echo: true
#| output-location: fragment
 library(psych)

describe(subdata)

```

## Descriptive Statistics

-   Without Skewness and Kurtosis

::: fragment
```{r}
#| echo: true
#| output-location: fragment
# without skewness and kurtosis
describe(subdata, skew=FALSE) 
```
:::

-   Without range

. . .

```{r}
#| echo: true
#| output-location: fragment
 # without range
describe(subdata, ranges = FALSE)
```

<!-- ## Decriptive statistics -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- #| output-location: fragment -->

<!-- # without quantiles -->

<!-- describe(subdata, quant = c(0.25,0.75)) -->

<!-- ``` -->

<!-- ## Decriptive statistics -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- #| output-location: fragment -->

<!-- # with quantiles -->

<!-- describe(subdata, quant = c(0.25,0.75), IQR=TRUE) -->

<!-- ``` -->

## Descriptive Statistics

-   Summary Statistics by grouping data using some specific criteria

. . .

```{r}
#| echo: true
#| output-location: fragment
# generating summary statistics by grouping variable
describeBy(subdata, datai$schtyp)
```

::: footer
To know more about `describe` and `describeBy` function use `help(describe)` and `?describeBy`
:::

# Frequencies and Cross Tabulation {background="#13464F"}

## Frequency Table

-   Generating Frequency Tables
-   frequency tables using the `table( )` function

. . .

```{r}
#| echo: true
#| output-location: fragment
table(datai$gender)
table(datai$schtyp)
```

## Frequency Tables

-   tables of proportions using the `prop.table( )` function
-   for proportions, use output of `table()` as input to `prop.table()`

. . .

```{r}
#| echo: true
#| output-location: fragment
#saving the freq table to an object
tableg <- table(datai$gender)
prop.table(tableg)
```

. . .

```{r}
#| echo: true
#| output-location: fragment
# OR
prop.table(table(datai$gender))
```

<!-- - marginal frequencies using margin.table( ) -->

<!-- . . .  -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- #| output-location: fragment -->

<!-- margin.table(tableg) -->

<!-- ``` -->

## Cross Tabulation {.scrollable}

-   Two Way Tabulation

-   counts in each crossing of gender and school type

. . .

```{r}
#| echo: true
#| output-location: fragment
tab2way <- table(datai$gender, datai$schtyp)
tab2way

```

-   Marginal frequencies using `margin.table( )`

. . .

```{r}
#| echo: true
#| output-location: fragment
margin.table(tab2way,margin = 1)

```

. . .

```{r}
#| echo: true
#| output-location: fragment
margin.table(tab2way,margin = 2)

```

## Proportion Table

-   Row proportions
-   Proportion of gender that falls into school type

. . .

```{r}
#| echo: true
#| output-location: fragment
prop.table(tab2way, margin = 1)

```

-   columns proportions,
-   Proportion of school type that falls into gender

. . .

```{r}
#| echo: true
#| output-location: fragment
prop.table(tab2way, margin = 2)

```

<!-- - Total Proportion  -->

<!-- ```{r} -->

<!-- #| echo: true -->

<!-- #| output-location: fragment -->

<!-- prop.table(tab2way) -->

<!-- ``` -->

# Correlation {background="#13464F"}

## Correlation {.scrollable}

-   Can use the `cor( )` function to produce correlations

-   General framework `cor(x, use=, method= )`

    -   `x`: Matrix or data frame
    -   `use`: Specifies the handling of missing data
    -   `method`: Specifies the type of correlation

. . .

```{r}
#| echo: true
#| output-location: fragment
cordata <- datai[,c("read","write","math")]
cor(cordata,use="all.obs",method="pearson")

```

. . .

```{r}
#| echo: true
#| output-location: fragment
cor(cordata,use="all.obs",method="spearman")

```

. . .

```{r}
#| echo: true
#| output-location: fragment
cor(cordata,use="all.obs",method="kendall")

```

-   `all.obs` assumes no missing data - missing data will produce an error
-   `complete.obs`-listwise deletion
-   `pairwise.complete.obs`- pairwise deletion

# THANKS {background-color="aquamarine"}
