---
title: "Exploring the Imported Data"
author: "Md Zulquar Nain"
format: 
  revealjs:
    incremental: true
    slide-number: c/t
    theme: dark
    preview-links: auto
    chalkboard: 
      boardmarker-width: 5
editor: visual
---

## Importing the Data File

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| evaluation: false
# importing data from `csv` file
datai <- read.csv("hsbraw.csv")
```

-   `datai` - name of the imported data file in`R`

-   `hsbraw.csv` name of the `csv` file being imported

## Exploring the Dataset I {background="#4682B4"}

-   Class, structure and dimension of the dataset

. . .

```{r }
#| echo: true
#| output-location: fragment

# Structure of the data
str(datai)
```

. . .

```{r }
#| echo: true
#| output-location: column-fragment


#Class of the data
class(datai)

```

. . .

```{r }
#| echo: true
#| output-location: column-fragment

# Dimension of the data
dim(datai)


```

## Exploring the Dataset II {background="#4682B4"}

-   First `n` rows of observations of the data set
    -   `head(data.frame name, n)`
-   Last `n` rows of observations of the data set
    -   `tail(data.frame name, n)`

. . .

```{r}
#| echo: true
#| output-location: fragment

# View top two rows of the data
head(datai,2)
```

. . .

```{r}
#| echo: true
#| output-location: fragment

# View bottom two rows 
tail(datai,2)
```

## Creating New Variables {background="#4682B4"}

-   Adding TWO variables & Creating a new ONE
-   Creating a new variable by taking average of two variables

. . .

```{r}
#| echo: true
#| output-location: fragment

ssum <- datai$read+datai$write
savg <- (datai$read+datai$write)/2

```

-   alternate method

. . .

```{r}
#| echo: true
#| output-location: fragment

attach(datai)
ssum1 <- read+write
savg1 <- (read+write)/2
detach(datai)

```

## Creating New Variables {background="#4682B4"}

-   Another method

. . .

```{r}
#| echo: true
#| output-location: fragment

dataim<- transform(datai,sum2=read+write,smean=(read+write)/2)

```

. . .

```{r}
#| echo: true
#| output-location: fragment
head(datai, 2)

```

. . .

```{r}
#| echo: true
#| output-location: fragment
head(dataim,2)

```


## Sorting/Ranking/Ordering {.scrollable}

-   Use `order()`

-   sort by a variable

. . .

```{r}
#| echo: true
#| output-location: fragment
max(dataim$write)
min(dataim$write)

```

. . .

```{r}
#| echo: true
#| output-location: fragment
head(dataim,4)

```

. . .

```{r}
#| echo: true
#| output-location: fragment
datais <- dataim[order(dataim$write),]

```

```{r}
#| echo: true
#| output-location: fragment
head(datais,4)

```

. . .

-   sort by descending order

. . .

```{r}
#| echo: true
#| output-location: fragment
dataisd <- dataim[order(-dataim$write),]
```

. . .

```{r}
#| echo: true
#| output-location: fragment
head(dataisd,4)

```


## Reshaping

-   Transpose

. . .

```{r}
#| echo: true
#| output-location: column-fragment

dataA <- data.frame(id=c(1,1,2,2),
                    time=c(1,2,1,2),
                    x1=c(5,2,4,6))
dim(dataA)
```

. . .

```{r}
#| echo: true
#| output-location: column-fragment
trdataA <- t(dataA)
dim(trdataA)
```


# THANKS {background-color="aquamarine"}
