---
title: "Exploratory Data Analysis"
author: "Zulquar Nain"
format: 
  revealjs:
    incremental: true
    scrollable: true
    slide-number: c/t
    theme: dark
    preview-links: auto
    chalkboard: 
      boardmarker-width: 5
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(quantmod)
library(tidyquant)
library(broom)
library(ggpubr)
library(zoo)
library(corrplot)
library(tseries)
library(car)
```

## What is EDA ?

-   Statistical approach of analyzing datasets and summarizing their main characteristics.

-   Statistics and visualization techniques used to explore data and discover patterns.

-   main idea is to better understand the data.

## Indian Stock Market - A Case Study

-   Indian sectoral Indices

-   The dataset comprises three sectoral indices from 2020 to 2025

. . .

### R packages required

```{r}
#| echo: true
#| output-location: fragment
# To attach the package
library(tidyverse)
library(quantmod)
library(tidyquant)
library(broom)
library(ggpubr)
library(zoo)
library(corrplot)
library(tseries)
library(car)
```

# Questions to Explore

1.  **What is the overall trend in sector indices from 2020 to 2025?**
2.  **Are daily returns normally distributed?**
3.  **Is there a significant difference between average returns of sectors?**
4.  **Are sector returns correlated?**
5.  **Can we predict Bank returns based on IT and FMCG returns?**

## 1. Import the data

```{r}
#| echo: true
#| output-location: fragment

# Define the list of sector symbols (from Yahoo Finance)
symbols <- c("^NSEBANK", "^CNXIT", "^CNXFMCG")

# Download data for each symbol from Yahoo Finance
# getSymbols() fetches the historical price data
# Here map() is for looping that is downloads stock data for each symbol and collects all the results into a list called data_list
# .x is just a placeholder. Typically, .x is used inside functions like purrr::map() or lapply(), where it represents each element of a list or vector you are looping over.

data_list <- map(
  symbols,
  ~ getSymbols(.x, src = "yahoo", from = "2020-01-01", 
               to = Sys.Date(), auto.assign = FALSE)
)

# Assign readable names to the downloaded data
# This matches the downloaded list to sector names
names(data_list) <- c("NIFTY_BANK", "NIFTY_IT", "NIFTY_FMCG")

# Convert each stock data into a tidy format
# Extract columns, rename them, add the Date and Sector Index
sector_tbl <- map2(data_list, names(data_list), ~ {
  
  # Convert xts object to tibble (dataframe)
  df <- as_tibble(as.data.frame(.x))
  
  # Rename the columns for clarity
  colnames(df) <- c("Open", "High", "Low", "Close", "Volume", "Adjusted")
  
  # Add the Date column (index of xts object)
  df <- df %>% mutate(Date = index(.x))
  
  # Add a new column "Index" to identify the sector
  df %>%
    mutate(Index = .y) %>%
    select(Date, Index, Close)   # Keep only relevant columns
})

# Combine all sectors into one dataframe
# Bind rows of all sector tibbles into a single dataframe
sector_df <- bind_rows(sector_tbl)

# Ensure the Date column is in Date format
sector_df$Date <- as.Date(sector_df$Date)

# View the structure of sector_df
glimpse(sector_df)
```

## 2. Overall Trend in Sector Indices

```{r}
#| echo: true
#| output-location: fragment

sector_df %>%
  ggplot(aes(x = Date, y = Close, color = Index)) +
  geom_line() +
  labs(title = "Sector Indices Closing Prices (2020-2025)", x = "Date",
       y = "Closing Price (INR)") +
  theme_minimal()
```

**Interpretation:** All sectors show upward trends with some volatility, especially Banking.

## 3. Calculating Daily Returns

```{r}
#| echo: true
#| output-location: fragment
sector_returns <- sector_df %>%
  group_by(Index) %>%
  arrange(Date) %>%
  mutate(Return = (Close / lag(Close)) - 1) %>%
  drop_na()
# View the structure of sector_df
glimpse(sector_returns)

```

## 4. Distribution of Daily Returns

```{r}
#| echo: true
#| output-location: fragment
sector_returns %>%
  ggplot(aes(x = Return, fill = Index)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  facet_wrap(~Index, scales = "free") +
  labs(title = "Histogram of Daily Returns", x = "Daily Return",
       y = "Frequency") +
  theme_minimal()
```

## 5. Normality Check with QQ Plot

```{r}
#| echo: true
#| output-location: fragment
sector_returns %>%
  ggqqplot(x = "Return", facet.by = "Index", color = "Index", add = "qqline")
```

**Interpretation:** Returns show deviation from perfect normality, common in financial data.

## 6. Hypothesis Testing - Are Average Returns Different?

```{r}
#| echo: true
#| output-location: fragment
# ANOVA Test
anova_test <- aov(Return ~ Index, data = sector_returns)
summary(anova_test)
```

**Interpretation:** If p-value \< 0.05, we reject the null hypothesis and conclude that mean returns differ across sectors.

## 7. Correlation Analysis

```{r}
#| echo: true
#| output-location: fragment
# Reshape returns to wide format
returns_wide <- sector_returns %>%
  select(Date, Index, Return) %>%
  pivot_wider(names_from = Index, values_from = Return)

# Correlation matrix
cor_matrix <- cor(returns_wide %>% select(-Date), use = "complete.obs")

# Correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.cex = 0.8, number.cex = 0.7,
         title = "Correlation Between Sector Returns", mar = c(0,0,1,0))
```

**Interpretation:** Positive correlations imply that sectors often move together. Banking and IT sectors show moderate correlation.

## 8. Predict Banking Returns using Regression Analysis

```{r}
#| echo: true
#| output-location: fragment
# Simple regression model
regression_model <- lm(NIFTY_BANK ~ NIFTY_IT + NIFTY_FMCG, data = returns_wide)
summary(regression_model)
```

**Interpretation:** - Coefficients tell us how much Bank returns change with IT and FMCG. - If p-values \< 0.05, predictors are significant. - Adjusted R-squared shows goodness of fit.

## 9. Final Inference

**Summary:** - All sectors have shown an overall upward trend. - Returns are not perfectly normal. - There are significant differences in sector returns. - Sector returns are moderately correlated. - IT and FMCG sector returns are useful predictors of Bank sector returns.

**Conclusion:** Investors can use sector correlations and predictive models for portfolio diversification and risk management.

# THANKS
