---
title: "Sectoral Analysis of Indian Stock Market Indices"
author: "Student Author Name"
format:
  docx:
    toc: true
    number-sections: true
    toc-location: body
    toc-title: "Table of Contents"
    toc-depth: 3
execute:
  echo: false
  warning: false
  message: false
crossref:
  labels: roman
  title-delim: "-"
#  enabled: true     # Required for numbering
  fig-prefix: "Fig." # Customize prefix
bibliography: references.bib
#csl: apa.csl
---

\newpage 

# Abstract{.unnumbered}

Abstract of the project goes here.

\newpage

# Introduction

Introduction of the study goes here. For example, see the below paragraph. <break>
Sectoral indices reflect the performance of key industries within the stock market. Understanding their behavior is crucial for investors, policymakers, and analysts. This report addresses the following research questions:

- What are the trends in sector indices from 2020–2025?
- Are daily returns normally distributed?
- Do average returns differ significantly across sectors?
- Are sector returns correlated?
- Can banking returns be predicted using IT and FMCG?
- Which sector is the most volatile?
- Do returns follow a random walk?
- Is there a causal relationship among sectors?

# Data and Methodology

```{r}
# Load required libraries
library(tidyverse)
library(quantmod)
library(tseries)
library(lmtest)
library(broom)
library(ggcorrplot)
library(urca)
library(vars)
library(knitr)
```

## Data 

Write here about the data here.

```{r}
#| label: return calculation
symbols <- c("^NSEBANK", "^CNXIT", "^CNXFMCG")

# Download historical data
data_list <- map(symbols, ~ getSymbols(.x, src = "yahoo", 
                                       from = "2020-01-01", 
                                       to = "2025-01-01", 
                                       auto.assign = FALSE))

# Name elements
names(data_list) <- c("Bank", "IT", "FMCG")

# Calculate returns and rename columns
returns_data <- imap(data_list, ~ {
  dailyReturn(Cl(.x)) %>%
    as.data.frame() %>%
    rownames_to_column(var = "Date") %>%
    rename(!!.y := 2)
})

# Merge all returns by Date
returns_df <- reduce(returns_data, full_join, by = "Date")
returns_df$Date <- as.Date(returns_df$Date)
```

## Methodology

Write here, about the methods used for the analysis.


# Results and discussion

Write here about the results which have been obtained after employing above stated methodologies and discuss about the results. 

## Trend Analysis

```{r}
#| label: Trend Analysis
data_list %>%
  imap(~ggplot(data = tibble(Date = index(.x), Close = Cl(.x))) +
         geom_line(aes(x = Date, y = Close), color = "blue") +
         labs(title = paste("Trend of", .y), x = "Date", y = "Closing Price") +
         theme_minimal())
```

## Normality of Returns

```{r}
#| label: Normality Check
returns_long <- returns_df %>%
  pivot_longer(cols = c(Bank, IT, FMCG), names_to = "Sector", values_to = "Return")

ggplot(returns_long, aes(x = Return)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  facet_wrap(~Sector, scales = "free") +
  theme_minimal()
```

## Difference in Mean Returns

```{r}
#| label: Mean Difference
aov_model <- aov(Return ~ Sector, data = returns_long)
kable(tidy(aov_model), caption = "Table 1: ANOVA Results for Sectoral Returns")
```

## Correlation Between Sectors

```{r }
#| label: Fig-Correlation Matrix
#| fig-cap: Correlation among sectorl returns
#| fig-align: center
#| fig-height: 5 
#| fig-width: 5 

cor_matrix <- returns_df %>%
  dplyr::select(Bank, IT, FMCG) %>%
  cor(use = "complete.obs")

ggcorrplot(cor_matrix, lab = TRUE)
```

## Predicting Banking Returns

```{r}
#| label: return prediction
lm_model <- lm(Bank ~ IT + FMCG, data = returns_df)
kable(tidy(lm_model), caption = "Table 2: Regression Model Predicting Bank Returns")
```

## Volatility Comparison

```{r}
#| label: Volatility 
volatility <- returns_df %>%
  summarise(across(c(Bank, IT, FMCG), ~ sd(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "Sector", values_to = "Volatility")
kable(volatility, caption = "Table 3: Sector-wise Volatility")
```



The analysis indicates that sectoral indices show distinct dynamics. Banking sector returns exhibit higher volatility, while FMCG remains relatively stable. ANOVA and regression models suggest statistically significant differences and predictive relationships across sectors. ADF results reject the random walk hypothesis, suggesting the market is not fully efficient.

# Conclusion

This report highlights structural differences across sectors, predictive linkages, and implications for market efficiency. These insights support diversification strategies and inform policy planning in financial regulation.

# References

- Fama, E. F. (1970). Efficient Capital Markets: A Review of Theory and Empirical Work.
- Lo, A. W., & MacKinlay, A. C. (1988). Stock Market Prices Do Not Follow Random Walks.
- Mishra, A. K., & Sethi, D. (2021). Sectoral Returns and Macroeconomic Variables: Evidence from Indian Stock Market.

# Appendix: R Code

```{r}
# Display full R code here 
# You can copy the codes used in the above analyses and write here with brief explanations.
```
