---
title: "India Inflation Tracker"
subtitle: "Retail inflation (CPI) pulled live from MoSPI's open API — trend, sectors, spending divisions, and contributions to the headline rate"
date: today
format:
  html:
    # NOT self-contained on purpose. embed-resources inlines DataTables,
    # ggiraph and every image into one ~11 MB file, and a fresh 11 MB blob
    # lands in git history at each monthly re-render. Split out, the JS
    # libraries in inflation-tracker_files/ stay byte-identical month to
    # month and git stores them once; only the charts churn.
    embed-resources: false
    toc: true
    toc-depth: 2
    toc-title: "On this page"
    fig-width: 9
    fig-height: 4.6
    css: tracker.css
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lubridate)
  library(httr2)
  library(jsonlite)
  library(stringr)
  library(forcats)
  library(scales)
})
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")

# ---- palette (validated reference palette; see notes) -----------------------
PAL <- list(
  surface = "#fcfcfb", ink = "#0b0b0b", ink2 = "#52514e", muted = "#898781",
  grid = "#e1e0d9", axis = "#c3c2b7", band = "#f0efec",
  blue = "#2a78d6", aqua = "#1baf7a", yellow = "#eda100", green = "#008300",
  violet = "#4a3aa7", red = "#e34948", magenta = "#e87ba4", orange = "#eb6834",
  down_good = "#006300", up_bad = "#d03b3b"
)

theme_tracker <- function(base_size = 12.5) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background = element_rect(fill = PAL$surface, colour = NA),
      panel.background = element_rect(fill = PAL$surface, colour = NA),
      panel.grid.major.y = element_line(colour = PAL$grid, linewidth = 0.35),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(colour = PAL$axis, linewidth = 0.4),
      axis.ticks = element_blank(),
      axis.text = element_text(colour = PAL$muted, size = rel(0.85)),
      axis.title = element_text(colour = PAL$ink2, size = rel(0.85)),
      plot.title = element_text(colour = PAL$ink, face = "bold", size = rel(1.12)),
      plot.subtitle = element_text(colour = PAL$ink2, size = rel(0.88), margin = margin(b = 10)),
      plot.caption = element_text(colour = PAL$muted, size = rel(0.72), hjust = 0, margin = margin(t = 10)),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      legend.position = "top",
      legend.justification = "left",
      legend.title = element_blank(),
      legend.text = element_text(colour = PAL$ink2, size = rel(0.85)),
      strip.text = element_text(colour = PAL$ink2, face = "bold", size = rel(0.88)),
      plot.margin = margin(8, 16, 8, 8)
    )
}

pct1 <- function(x) ifelse(is.na(x), "—", sprintf("%.1f%%", x))
pp1  <- function(x) ifelse(is.na(x), "—", sprintf("%+.1f pp", x))

# ---- MoSPI API client -------------------------------------------------------
MOSPI_BASE <- "https://api.mospi.gov.in"
CACHE_DIR  <- "data_cache"
dir.create(CACHE_DIR, showWarnings = FALSE)

mospi_get <- function(endpoint, params, limit = 100, max_pages = 250) {
  pages <- list()
  page  <- 1L
  repeat {
    resp <- request(MOSPI_BASE) |>
      req_url_path(endpoint) |>
      req_url_query(!!!params, Format = "JSON", limit = limit, page = page) |>
      req_user_agent("india-inflation-tracker (R/httr2)") |>
      req_timeout(120) |>
      req_retry(max_tries = 3, backoff = ~2) |>
      req_perform()
    dat <- fromJSON(resp_body_string(resp))$data
    if (is.null(dat) || NROW(dat) == 0) break
    pages[[page]] <- as_tibble(dat)
    if (NROW(dat) < limit) break
    page <- page + 1L
    if (page > max_pages) break
  }
  bind_rows(pages)
}

# Fetch with a local cache. On API failure a stale cache is used silently;
# a missing cache plus a failed fetch returns an empty tibble.
cached_fetch <- function(file, max_age_hours, fetch_fn) {
  path <- file.path(CACHE_DIR, file)
  if (file.exists(path)) {
    age_h <- as.numeric(difftime(Sys.time(), file.mtime(path), units = "hours"))
    if (age_h < max_age_hours) return(readRDS(path))
  }
  fresh <- tryCatch(fetch_fn(), error = function(e) NULL)
  if (!is.null(fresh) && NROW(fresh) > 0) {
    saveRDS(fresh, path)
    return(fresh)
  }
  if (file.exists(path)) readRDS(path) else tibble()
}

month_date <- function(year, month_name) {
  as.Date(sprintf("%s-%02d-01", year, match(month_name, month.name)))
}
```

```{r fetch}
#| cache: false
this_year <- year(Sys.Date())

# Old series, base 2012 = 100 (Jan 2013 - Dec 2025, discontinued). Refreshed monthly
# in case of revisions; delete data_cache/ to force a full re-download.
d12_raw <- cached_fetch(
  "cpi_base2012_allindia.rds", max_age_hours = 24 * 30,
  function() mospi_get(
    "/api/cpi/getCPIIndex",
    params = list(base_year = "2012", series = "Current", state_code = "99"),
    limit = 2000
  )
)

# New series, base 2024 = 100 (2025 onwards). Combined sector, full division
# hierarchy, one cache file per year; the current year expires after 12 hours.
d24_raw <- bind_rows(lapply(2025:this_year, function(y) {
  cached_fetch(
    sprintf("cpi_base2024_combined_%d.rds", y),
    max_age_hours = if (y < this_year) 24 * 30 else 12,
    function() mospi_get(
      "/api/cpi/getCPIData",
      params = list(base_year = "2024", series = "Current",
                    state_code = "1", sector_code = "3", year = y),
      limit = 100
    )
  )
}))

# New series, headline (division 0) only, for all three sectors.
d24_gen_raw <- bind_rows(lapply(2025:this_year, function(y) {
  cached_fetch(
    sprintf("cpi_base2024_general_%d.rds", y),
    max_age_hours = if (y < this_year) 24 * 30 else 12,
    function() mospi_get(
      "/api/cpi/getCPIData",
      params = list(base_year = "2024", series = "Current",
                    state_code = "1", division_code = "0", year = y),
      limit = 100
    )
  )
}))

# New series, headline (division 0) by state/UT, Combined sector. Omitting
# state_code returns every state plus All India. Optional: the map section
# is skipped gracefully if this comes back empty.
dst_raw <- bind_rows(lapply(2025:this_year, function(y) {
  cached_fetch(
    sprintf("cpi_base2024_states_%d.rds", y),
    max_age_hours = if (y < this_year) 24 * 30 else 12,
    function() mospi_get(
      "/api/cpi/getCPIData",
      params = list(base_year = "2024", series = "Current",
                    sector_code = "3", division_code = "0", year = y),
      limit = 100
    )
  )
}))

if (NROW(d12_raw) == 0 || NROW(d24_raw) == 0 || NROW(d24_gen_raw) == 0) {
  stop(paste(
    "Could not reach the MoSPI API (api.mospi.gov.in) and no local cache exists yet.",
    "Check the connection and re-render; the first successful render caches everything."
  ))
}
```

```{r wrangle}
# ---- official weighting diagrams (MoSPI press release, 12 Feb 2026) ---------
divisions <- tribble(
  ~division,                                                                 ~short,               ~weight,
  "Food and beverages",                                                      "Food & beverages",     36.753,
  "Housing, water, electricity, gas and other fuels",                        "Housing & utilities",  17.665,
  "Transport",                                                               "Transport",             8.796,
  "Clothing and footwear",                                                   "Clothing & footwear",   6.383,
  "Health",                                                                  "Health",                6.100,
  "Personal care, social protection and miscellaneous goods and services",   "Personal care & misc",  5.038,
  "Furnishings, household equipment and routine household maintenance",      "Household goods",       4.469,
  "Information and communication",                                           "Info & communication",  3.609,
  "Restaurants and accommodation services",                                  "Restaurants & hotels",  3.348,
  "Education services",                                                      "Education",             3.333,
  "Paan, tobacco and intoxicants",                                           "Paan & tobacco",        2.989,
  "Recreation, sport and culture",                                           "Recreation & culture",  1.516
)

# ---- old series (2012 base): group-level, All India -------------------------
d12 <- d12_raw |>
  mutate(
    date = month_date(year, month),
    index = suppressWarnings(as.numeric(index)),
    yoy = suppressWarnings(as.numeric(inflation))
  ) |>
  filter(!is.na(date))

d12_group <- d12 |>
  filter(is.na(subgroup) | subgroup == "" | subgroup == group |
           str_ends(subgroup, "-Overall"))

headline12 <- d12_group |>
  filter(group == "General", sector == "Combined", !is.na(yoy)) |>
  distinct(date, .keep_all = TRUE) |>
  arrange(date)

# ---- new series (2024 base): division-level, All India ----------------------
d24 <- d24_raw |>
  mutate(
    date = month_date(year, month),
    index = suppressWarnings(as.numeric(index)),
    yoy = suppressWarnings(as.numeric(inflation))
  ) |>
  filter(!is.na(date))

d24_div <- d24 |>
  filter(is.na(group), is.na(item)) |>            # division-level rows only
  distinct(date, sector, division, .keep_all = TRUE)

d24_grp <- d24 |>
  filter(!is.na(group), is.na(`class`), is.na(item)) |>   # 43 COICOP groups
  distinct(date, sector, division, group, .keep_all = TRUE)

headline24 <- d24_gen_raw |>
  mutate(
    date = month_date(year, month),
    index = suppressWarnings(as.numeric(index)),
    yoy = suppressWarnings(as.numeric(inflation))
  ) |>
  filter(division == "CPI (General)", !is.na(date)) |>
  distinct(date, sector, .keep_all = TRUE)

# ---- chained headline: 2012 base through Dec 2025, 2024 base from Jan 2026 --
SWITCH_DATE <- as.Date("2026-01-01")

headline <- bind_rows(
  headline12 |> filter(date < SWITCH_DATE) |>
    transmute(date, yoy, base = "2012 = 100"),
  headline24 |> filter(sector == "Combined", date >= SWITCH_DATE, !is.na(yoy)) |>
    transmute(date, yoy, base = "2024 = 100")
) |> arrange(date)

latest_date  <- max(headline$date)
latest_month <- format(latest_date, "%B %Y")
latest       <- headline |> filter(date == latest_date)
prev         <- headline |> filter(date == latest_date %m-% months(1))

# month-on-month, computed within the continuous 2024-base index
mom_series <- headline24 |>
  filter(sector == "Combined") |>
  arrange(date) |>
  mutate(mom = 100 * (index / lag(index) - 1))
latest_mom <- mom_series |> filter(date == latest_date) |> pull(mom)

latest_div <- d24_div |>
  filter(sector == "Combined", date == latest_date, division != "CPI (General)") |>
  inner_join(divisions, by = "division")

food_yoy <- latest_div |> filter(short == "Food & beverages") |> pull(yoy)
food_prev <- d24_div |>
  filter(sector == "Combined", date == latest_date %m-% months(1),
         division == "Food and beverages") |> pull(yoy)

# ---- state-level headline: latest month available in the state data ---------
dstate <- dst_raw |>
  mutate(
    date = month_date(year, month),
    yoy = suppressWarnings(as.numeric(inflation))
  ) |>
  filter(state != "All India", division == "CPI (General)",
         !is.na(date), !is.na(yoy)) |>
  distinct(date, state, .keep_all = TRUE)

state_date <- if (NROW(dstate) > 0) max(dstate$date) else as.Date(NA)

# colour scale shared by the division heatmap and the state map:
# saturates 6 pp either side of the 4% target
HEAT_SPAN <- 6

sect_latest <- headline24 |> filter(date == latest_date) |> select(sector, yoy)
rural_yoy <- sect_latest |> filter(sector == "Rural") |> pull(yoy)
urban_yoy <- sect_latest |> filter(sector == "Urban") |> pull(yoy)

# ---- contribution of each division to headline YoY --------------------------
# index a year ago is recovered from the published YoY rate, so contributions
# use only official numbers: contrib_g = w_g * (I_t - I_t-12) / sum(w * I_t-12)
contrib <- d24_div |>
  filter(sector == "Combined", division != "CPI (General)", !is.na(yoy)) |>
  inner_join(divisions, by = "division") |>
  mutate(idx_prev = index / (1 + yoy / 100)) |>
  group_by(date) |>
  mutate(contrib = weight * (index - idx_prev) / sum(weight * idx_prev) * 100) |>
  ungroup()

contrib_latest <- contrib |> filter(date == latest_date)

# ---- status file ------------------------------------------------------------
# Written on every render and read by two other things: print_software() in
# R/helpers.R stamps "Data through ..." on the Software page card, and the
# check-inflation-data GitHub Action compares latest_month against what MoSPI
# is publishing and opens an issue when this page falls behind. Keep the
# field names in step with both if you change them.
writeLines(
  jsonlite::toJSON(
    list(
      latest_month = format(latest_date, "%Y-%m"),
      latest_label = format(latest_date, "%b %Y"),
      headline_yoy = round(latest$yoy, 2),
      rendered     = format(Sys.Date())
    ),
    auto_unbox = TRUE, pretty = TRUE
  ),
  "tracker_status.json"
)
```

```{r kpis, results='asis'}
delta_html <- function(delta, unit = "pp") {
  if (length(delta) == 0 || is.na(delta)) return("")
  cls <- if (delta <= 0) "kpi-delta down" else "kpi-delta up"
  arw <- if (delta <= 0) "&#9660;" else "&#9650;"
  sprintf('<span class="%s">%s %+.2f %s vs prior month</span>', cls, arw, delta, unit)
}
tile <- function(label, value, sub) {
  sprintf('<div class="kpi"><div class="kpi-label">%s</div><div class="kpi-value">%s</div>%s</div>',
          label, value, sub)
}
cat(
  '<div class="kpi-grid">',
  tile("Headline inflation (y/y)", pct1(latest$yoy),
       delta_html(latest$yoy - prev$yoy)),
  tile("Food &amp; beverages (y/y)", pct1(food_yoy),
       delta_html(food_yoy - food_prev)),
  tile("Headline, month-on-month", pct1(latest_mom),
       sprintf('<span class="kpi-sub">seasonally unadjusted</span>')),
  tile("Rural &nbsp;|&nbsp; Urban (y/y)",
       paste0(pct1(rural_yoy), " | ", pct1(urban_yoy)),
       sprintf('<span class="kpi-sub">gap %s</span>', pp1(rural_yoy - urban_yoy))),
  '</div>',
  sprintf('<p class="kpi-note">Latest data: <strong>%s</strong>, All-India CPI, base 2024&nbsp;=&nbsp;100 · fetched %s from api.mospi.gov.in</p>',
          latest_month, format(Sys.time(), "%d %b %Y %H:%M")),
  sep = "\n"
)
```

## Headline trend

Consumer price inflation in `r latest_month` was **`r pct1(latest$yoy)`** year on year.
The chart chains the discontinued 2012-base series (through December 2025) with the
new 2024-base series that MoSPI introduced in February 2026; the dotted marker shows
the switch. The shaded band is the RBI's flexible inflation-targeting range of
4% ± 2 percentage points.

```{r fig-trend}
#| fig-height: 4.8
#| fig-alt: "Line chart of India headline CPI inflation, year on year, from 2014 to the latest month, with the RBI 2 to 6 percent tolerance band shaded and the 4 percent target marked."
trend <- headline |> filter(date >= as.Date("2014-01-01"))

ggplot(trend, aes(date, yoy)) +
  annotate("rect", xmin = min(trend$date), xmax = max(trend$date) %m+% months(4),
           ymin = 2, ymax = 6, fill = PAL$band) +
  geom_hline(yintercept = 4, colour = PAL$muted, linetype = "42", linewidth = 0.4) +
  geom_vline(xintercept = SWITCH_DATE, colour = PAL$muted, linetype = "13", linewidth = 0.45) +
  geom_line(colour = PAL$blue, linewidth = 0.85) +
  geom_point(data = slice_tail(trend, n = 1), colour = PAL$blue, size = 2.4) +
  geom_text(data = slice_tail(trend, n = 1),
            aes(label = sprintf("%.1f%%", yoy)),
            hjust = -0.35, vjust = 0.5, colour = PAL$ink, fontface = "bold", size = 3.4) +
  annotate("text", x = SWITCH_DATE, y = max(trend$yoy, na.rm = TRUE),
           label = "new 2024 = 100 series", hjust = 1.05, vjust = 1,
           colour = PAL$muted, size = 3) +
  annotate("text", x = min(trend$date), y = 6, label = "RBI tolerance band 2–6%",
           hjust = -0.02, vjust = -0.6, colour = PAL$muted, size = 3) +
  scale_x_date(breaks = seq(as.Date("2014-01-01"), max(trend$date), by = "2 years"),
               date_labels = "%Y",
               expand = expansion(mult = c(0.01, 0.06))) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Headline retail inflation, year on year",
    subtitle = "All-India CPI (Combined) · 2012-base series to Dec 2025, 2024-base series from Jan 2026",
    x = NULL, y = NULL,
    caption = "Source: MoSPI CPI API (api.mospi.gov.in). YoY rates as published by MoSPI for each base series."
  ) +
  theme_tracker()
```

## Rural vs urban

```{r fig-sector}
#| fig-height: 4.4
#| fig-alt: "Line chart comparing rural and urban year-on-year CPI inflation over the last four years, with direct labels on the latest values."
sect <- bind_rows(
  d12_group |>
    filter(group == "General", sector %in% c("Rural", "Urban"),
           date < SWITCH_DATE, !is.na(yoy)) |>
    select(date, sector, yoy),
  headline24 |>
    filter(sector %in% c("Rural", "Urban"), date >= SWITCH_DATE, !is.na(yoy)) |>
    select(date, sector, yoy)
) |>
  filter(date >= latest_date %m-% years(4)) |>
  arrange(date)

sect_cols <- c(Rural = PAL$blue, Urban = PAL$aqua)
sect_ends <- sect |> group_by(sector) |> slice_max(date, n = 1) |> ungroup()

ggplot(sect, aes(date, yoy, colour = sector)) +
  geom_hline(yintercept = 0, colour = PAL$axis, linewidth = 0.4) +
  geom_vline(xintercept = SWITCH_DATE, colour = PAL$muted, linetype = "13", linewidth = 0.45) +
  geom_line(linewidth = 0.85) +
  geom_point(data = sect_ends, size = 2.2) +
  geom_text(data = sect_ends,
            aes(label = sprintf("%s %.1f%%", sector, yoy)),
            hjust = -0.12, vjust = 0.5, size = 3.3, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = sect_cols) +
  scale_x_date(breaks = seq(ceiling_date(min(sect$date), "year"), max(sect$date), by = "1 year"),
               date_labels = "%Y",
               expand = expansion(mult = c(0.01, 0.20))) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Rural and urban inflation, year on year",
    subtitle = "All-India CPI (General) by sector · last four years",
    x = NULL, y = NULL,
    caption = "Source: MoSPI CPI API. Rural housing enters the index only from the 2024-base series."
  ) +
  theme_tracker()
```

## Inflation across states

How the latest headline rate varies across states and union territories —
hover over a state to see its name and rate. Colours use the same scale as
the division heatmap below: white at the 4% target, red above, blue below.
Union territories too small to see at this scale — Delhi, Chandigarh,
Puducherry, Dadra & Nagar Haveli and Daman & Diu, and Lakshadweep — are
marked with dots.

```{r fig-statemap}
# Boundaries as published by the Government of India (Survey of India view:
# Jammu & Kashmir and Ladakh shown in full). District file dissolved to
# state level once, then cached alongside the API pulls.
GEO_URLS <- c(
  "https://raw.githubusercontent.com/udit-001/india-maps-data/main/geojson/india.geojson",
  "https://cdn.jsdelivr.net/gh/udit-001/india-maps-data@main/geojson/india.geojson"
)

india_sf <- local({
  path <- file.path(CACHE_DIR, "india_states_sf.rds")
  if (file.exists(path)) return(readRDS(path))
  g <- NULL
  for (u in GEO_URLS) {
    g <- tryCatch(sf::st_read(u, quiet = TRUE), error = function(e) NULL)
    if (!is.null(g)) break
  }
  if (is.null(g)) return(NULL)
  g <- g |> group_by(st_nm) |> summarise(.groups = "drop")
  saveRDS(g, path)
  g
})

# harmonise MoSPI state names with the map's ("NCT of Delhi" -> "Delhi",
# "The Dadra And ..." -> "Dadra and ...", "&" vs "and", casing)
norm_state <- function(x) {
  x <- str_to_lower(x)
  x <- str_replace_all(x, "&", "and")
  x <- str_remove(x, "^the ")
  x <- str_remove(x, "^nct of ")
  str_squish(x)
}

if (!is.null(india_sf) && NROW(dstate) > 0) {
  # ggiraph's sf geom is incompatible with ggplot2 4.x, so project to an
  # India-centric CRS, simplify on the plane (lighter SVG, type-stable),
  # and draw plain interactive polygons from the flattened coordinates.
  # (sf must act on the object before any dplyr verb so that its
  # data-frame methods are registered when the cache skips st_read.)
  smap_proj <- india_sf |>
    sf::st_transform(7755) |>
    sf::st_simplify(dTolerance = 2000, preserveTopology = TRUE) |>
    sf::st_cast("MULTIPOLYGON") |>
    mutate(key = norm_state(st_nm)) |>
    left_join(
      dstate |> filter(date == state_date) |>
        transmute(key = norm_state(state), yoy),
      by = "key"
    ) |>
    mutate(tip = ifelse(is.na(yoy),
                        sprintf("%s — no data", st_nm),
                        sprintf("%s — %.1f%% y/y", st_nm, yoy)))

  poly_df <- as_tibble(sf::st_coordinates(smap_proj)) |>
    transmute(x = X, y = Y, feat = L3,
              poly = paste(L3, L2, sep = "."),
              ring = interaction(L3, L2, L1)) |>
    left_join(
      smap_proj |> sf::st_drop_geometry() |>
        mutate(feat = row_number()) |>
        select(feat, st_nm, yoy, tip),
      by = "feat"
    )

  # print the rate on states large enough to hold a label
  lab_pts <- smap_proj |>
    filter(!is.na(yoy)) |>
    mutate(area_km2 = as.numeric(sf::st_area(geometry)) / 1e6) |>
    filter(area_km2 > 15000) |>
    sf::st_point_on_surface()
  lab_df <- bind_cols(
    lab_pts |> sf::st_drop_geometry() |> select(yoy),
    as_tibble(sf::st_coordinates(lab_pts))
  )

  # UTs only a few pixels wide at this scale (Delhi, Chandigarh, Puducherry,
  # DNH & Daman and Diu, Lakshadweep) get a hoverable dot so they stay visible
  small_pts <- smap_proj |>
    mutate(area_km2 = as.numeric(sf::st_area(geometry)) / 1e6) |>
    filter(area_km2 < 2000) |>
    sf::st_point_on_surface()
  small_df <- bind_cols(
    small_pts |> sf::st_drop_geometry() |> select(st_nm, yoy, tip),
    as_tibble(sf::st_coordinates(small_pts))
  )

  p <- ggplot(poly_df) +
    ggiraph::geom_polygon_interactive(
      aes(x, y, group = poly, subgroup = ring, fill = yoy,
          tooltip = tip, data_id = st_nm),
      colour = PAL$surface, linewidth = 0.35
    ) +
    ggiraph::geom_point_interactive(
      data = small_df,
      aes(X, Y, fill = yoy, tooltip = tip, data_id = st_nm),
      shape = 21, size = 3, colour = PAL$ink2, stroke = 0.5,
      inherit.aes = FALSE
    ) +
    geom_text(data = lab_df,
              aes(X, Y, label = sprintf("%.1f", yoy),
                  colour = abs(yoy - 4) > 0.72 * HEAT_SPAN),
              size = 2.8) +
    scale_fill_gradient2(low = PAL$blue, mid = PAL$band, high = PAL$red,
                         midpoint = 4, limits = c(4 - HEAT_SPAN, 4 + HEAT_SPAN),
                         oob = squish, breaks = c(-2, 0, 2, 4, 6, 8, 10),
                         na.value = PAL$grid, name = "y/y %") +
    scale_colour_manual(values = c(`FALSE` = PAL$ink2, `TRUE` = "#ffffff"),
                        guide = "none") +
    coord_fixed(expand = FALSE) +
    labs(
      title = sprintf("Headline inflation by state, %s", format(state_date, "%B %Y")),
      subtitle = "Year-on-year % change, CPI (General), Combined · white = at the 4% target, colour capped ±6 pp",
      x = NULL, y = NULL,
      caption = "Source: MoSPI CPI API, 2024-base series. Map: boundaries as published by the Government of India\n(Survey of India); state polygons dissolved from the district file of udit-001/india-maps-data."
    ) +
    theme_tracker() +
    theme(panel.grid = element_blank(),
          axis.line.x = element_blank(),
          axis.text = element_blank(),
          legend.key.height = unit(10, "pt"), legend.key.width = unit(28, "pt"))

  ggiraph::girafe(
    ggobj = p, width_svg = 9, height_svg = 8.4,
    options = list(
      ggiraph::opts_hover(css = "stroke:#0b0b0b; stroke-width:1.1px; cursor:pointer;"),
      ggiraph::opts_tooltip(
        css = paste0("background:", PAL$ink, "; color:#fff; padding:6px 10px;",
                     "border-radius:6px; font-size:0.85rem;"),
        opacity = 0.95
      ),
      ggiraph::opts_toolbar(saveaspng = FALSE),
      ggiraph::opts_sizing(rescale = TRUE, width = 1)
    )
  )
} else {
  cat("State-level data or the map file could not be fetched on this render;",
      "the state map will return on the next successful refresh.")
}
```

```{r tbl-states, results='asis'}
if (NROW(dstate) > 0) {
  st_tbl <- dstate |>
    filter(date == state_date) |>
    transmute(`State / UT` = state, `y/y %` = yoy)
  hi <- st_tbl |> slice_max(`y/y %`, n = 5)
  lo <- st_tbl |> slice_min(`y/y %`, n = 5) |> arrange(`y/y %`)
  cat("\n:::: {.movers}\n\n::: {}\n**Highest inflation**\n\n")
  print(knitr::kable(hi, digits = 1))
  cat("\n:::\n\n::: {}\n**Lowest inflation**\n\n")
  print(knitr::kable(lo, digits = 1))
  cat("\n:::\n\n::::\n")
}
```

## Sector-wise inflation

The new CPI classifies spending into twelve COICOP divisions. Bars show the latest
year-on-year rate for each division; the heatmap tracks how each rate has evolved
month by month under the new series, coloured by distance from the 4% target.

```{r fig-divbar}
#| fig-height: 5.4
#| fig-alt: "Horizontal bar chart of year-on-year inflation by CPI division for the latest month, sorted from highest to lowest, with negative values in red."
bar <- latest_div |>
  filter(!is.na(yoy)) |>
  mutate(short = fct_reorder(factor(short), yoy))

ggplot(bar, aes(yoy, short, fill = yoy < 0)) +
  geom_vline(xintercept = 0, colour = PAL$axis, linewidth = 0.4) +
  geom_col(width = 0.62) +
  geom_text(aes(label = sprintf("%.1f", yoy),
                hjust = ifelse(yoy < 0, 1.25, -0.25)),
            colour = PAL$ink2, size = 3.2) +
  scale_fill_manual(values = c(`FALSE` = PAL$blue, `TRUE` = PAL$red), guide = "none") +
  scale_x_continuous(labels = label_number(suffix = "%"),
                     expand = expansion(mult = c(0.08, 0.12))) +
  labs(
    title = sprintf("Inflation by division, %s", latest_month),
    subtitle = "Year-on-year % change, All-India Combined · CPI 2024 = 100",
    x = NULL, y = NULL,
    caption = "Source: MoSPI CPI API. Division names abbreviated; full names in the table below."
  ) +
  theme_tracker() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = PAL$grid, linewidth = 0.35),
        axis.line.x = element_blank(),
        axis.text.y = element_text(colour = PAL$ink2))
```

```{r fig-heat}
#| fig-height: 5.2
#| fig-alt: "Heatmap of year-on-year inflation by CPI division and month under the 2024-base series. Cells are blue below the 4 percent target and red above it, with the rate printed in each cell."
heat <- d24_div |>
  filter(sector == "Combined", division != "CPI (General)", !is.na(yoy)) |>
  inner_join(divisions, by = "division") |>
  mutate(short = factor(short, levels = rev(divisions$short)),
         mlab = factor(format(date, "%b %y"),
                       levels = format(sort(unique(date)), "%b %y")))

ggplot(heat, aes(mlab, short, fill = yoy)) +
  geom_tile(colour = PAL$surface, linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.1f", yoy),
                colour = abs(yoy - 4) > 0.72 * HEAT_SPAN),
            size = 2.9) +
  scale_fill_gradient2(low = PAL$blue, mid = PAL$band, high = PAL$red,
                       midpoint = 4, limits = c(4 - HEAT_SPAN, 4 + HEAT_SPAN),
                       oob = squish, breaks = c(-2, 0, 2, 4, 6, 8, 10),
                       name = "y/y %") +
  scale_colour_manual(values = c(`FALSE` = PAL$ink2, `TRUE` = "#ffffff"), guide = "none") +
  scale_x_discrete(position = "top") +
  labs(
    title = "Division inflation, month by month",
    subtitle = "Year-on-year % change · divisions ordered by CPI weight · white = at the 4% target, colour capped ±6 pp",
    x = NULL, y = NULL,
    caption = "Source: MoSPI CPI API, 2024-base series (published from Jan 2026; back months where available)."
  ) +
  theme_tracker() +
  theme(panel.grid.major.y = element_blank(),
        axis.line.x = element_blank(),
        axis.text.y = element_text(colour = PAL$ink2),
        legend.key.height = unit(10, "pt"), legend.key.width = unit(28, "pt"))
```

## What is driving headline inflation

Each division's pull on the headline rate is its inflation weighted by its share of
the consumption basket. Food & beverages carries 36.8% of the basket, so even modest
food inflation moves the total. Contributions are derived from published division
indices, published year-on-year rates, and the official weighting diagram.

```{r fig-contrib}
#| fig-height: 4.8
#| fig-alt: "Stacked monthly bars of each division's contribution to headline year-on-year inflation in percentage points, for the months of the 2024-base series, with the headline rate marked on each month."
top5 <- contrib |>
  group_by(short) |>
  summarise(m = mean(abs(contrib)), .groups = "drop") |>
  slice_max(m, n = 5) |>
  pull(short)

contrib_plot <- contrib |>
  mutate(grp = ifelse(short %in% top5, short, "All other divisions")) |>
  group_by(date, grp) |>
  summarise(contrib = sum(contrib), .groups = "drop") |>
  mutate(
    grp = factor(grp, levels = c(
      divisions$short[divisions$short %in% top5], "All other divisions")),
    mlab = factor(format(date, "%b %y"),
                  levels = format(sort(unique(date)), "%b %y"))
  )

head_pts <- contrib |>
  group_by(date) |>
  summarise(total = sum(contrib), .groups = "drop") |>
  mutate(mlab = factor(format(date, "%b %y"), levels = levels(contrib_plot$mlab)))

contrib_cols <- setNames(
  c(PAL$blue, PAL$aqua, PAL$yellow, PAL$green, PAL$violet, PAL$grid),
  levels(contrib_plot$grp)
)

ggplot(contrib_plot, aes(mlab, contrib, fill = grp)) +
  geom_hline(yintercept = 0, colour = PAL$axis, linewidth = 0.4) +
  geom_col(width = 0.6, colour = PAL$surface, linewidth = 0.4) +
  geom_point(data = head_pts, aes(mlab, total), inherit.aes = FALSE,
             colour = PAL$ink, size = 2.2) +
  geom_text(data = slice_tail(head_pts, n = 1),
            aes(mlab, total, label = sprintf("headline %.1f%%", total)),
            inherit.aes = FALSE, vjust = -1.2, hjust = 0.85,
            colour = PAL$ink, size = 3.2, fontface = "bold") +
  scale_fill_manual(values = contrib_cols) +
  scale_y_continuous(labels = label_number(suffix = " pp"),
                     expand = expansion(mult = c(0.02, 0.14))) +
  labs(
    title = "Contribution to headline inflation, by division",
    subtitle = "Percentage points of the year-on-year rate · All-India Combined · dots mark the reconstructed headline",
    x = NULL, y = NULL,
    caption = "Derived from MoSPI division indices, published YoY rates and the official 2024 weighting diagram.\nContributions sum to the headline rate up to rounding in the published indices."
  ) +
  theme_tracker()
```

```{r tbl-divisions}
#| fig-alt: "Sortable table of the latest month's CPI divisions."
tbl <- latest_div |>
  left_join(contrib_latest |> select(division, contrib), by = "division") |>
  arrange(desc(contrib)) |>
  transmute(
    Division = division,
    `Weight (%)` = weight,
    Index = index,
    `Inflation y/y (%)` = yoy,
    `Contribution (pp)` = round(contrib, 2),
    `Share of headline (%)` = round(100 * contrib / sum(contrib, na.rm = TRUE), 1)
  )

DT::datatable(
  tbl,
  rownames = FALSE,
  caption = sprintf("CPI divisions, %s — All-India Combined, base 2024 = 100", latest_month),
  options = list(paging = FALSE, searching = FALSE, info = FALSE,
                 order = list(list(4, "desc")))
) |>
  DT::formatRound(c("Weight (%)", "Index", "Inflation y/y (%)"), 1)
```

### Largest moves within the 43 COICOP groups

```{r tbl-movers, results='asis'}
movers <- d24_grp |>
  filter(sector == "Combined", date == latest_date, !is.na(yoy)) |>
  left_join(divisions |> select(division, short), by = "division") |>
  select(Group = group, Division = short, `y/y %` = yoy)

up <- movers |> slice_max(`y/y %`, n = 5)
dn <- movers |> slice_min(`y/y %`, n = 5)

cat("\n:::: {.movers}\n\n::: {}\n**Fastest rising**\n\n")
print(knitr::kable(up, digits = 1))
cat("\n:::\n\n::: {}\n**Falling / slowest**\n\n")
print(knitr::kable(dn, digits = 1))
cat("\n:::\n\n::::\n")
```

## The long view, by spending group (2012-base series)

The discontinued 2012-base series still provides twelve years of comparable
group-level history — useful context for how unusual (or not) the current
readings are. The grey line in each panel is headline inflation.

```{r fig-history}
#| fig-height: 6.6
#| fig-alt: "Small-multiple line charts of year-on-year inflation for the six CPI groups of the 2012-base series from 2014 to 2025, each compared against headline inflation in grey."
groups12 <- c("Food and Beverages", "Pan, Tobacco and Intoxicants",
              "Clothing and Footwear", "Housing", "Fuel and Light", "Miscellaneous")

hist_grp <- d12_group |>
  filter(sector == "Combined", group %in% groups12,
         date >= as.Date("2014-01-01"), !is.na(yoy)) |>
  mutate(group = factor(group, levels = groups12))

hist_ctx <- crossing(group = factor(groups12, levels = groups12),
                     headline12 |> filter(date >= as.Date("2014-01-01")) |>
                       select(date, hl = yoy))

ggplot(hist_grp, aes(date, yoy)) +
  geom_hline(yintercept = 0, colour = PAL$axis, linewidth = 0.35) +
  geom_line(data = hist_ctx, aes(date, hl), colour = PAL$axis, linewidth = 0.5) +
  geom_line(colour = PAL$blue, linewidth = 0.65) +
  facet_wrap(~group, ncol = 2) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Group inflation vs headline (grey), 2014–2025",
    subtitle = "Year-on-year % change, All-India Combined · CPI 2012 = 100 (series discontinued Dec 2025)",
    x = NULL, y = NULL,
    caption = "Source: MoSPI CPI API. Rural housing was outside the index before the 2024-base series."
  ) +
  theme_tracker()
```

## Data & method notes

- **Source.** All figures are fetched at render time from the Ministry of Statistics
  and Programme Implementations
  <!-- 's open API (`api.mospi.gov.in`) — the same backend -->
  as reported on the [eSankhyiki portal](https://esankhyiki.mospi.gov.in). 
  <!-- No API key is required. -->
  <!-- Endpoints: `getCPIIndex` (base 2012 = 100) and `getCPIData` (base 2024 = 100). -->
- **Base switch.** MoSPI moved the CPI to base 2024 = 100 (COICOP 2018, 12 divisions)
  with the January 2026 release. This page uses the 2012-base series through
  December 2025 and the 2024-base series from January 2026; year-on-year rates are
  as published for each series, so the chained trend matches official releases.
- **Weights** are the official weighting diagram from MoSPI's first 2024-base press
  release (12 Feb 2026): Food & beverages 36.75, Housing & utilities 17.67,
  Transport 8.80, Clothing 6.38, Health 6.10, Personal care & misc 5.04, Household
  goods 4.47, Info & communication 3.61, Restaurants 3.35, Education 3.33,
  Paan & tobacco 2.99, Recreation 1.52 (All-India Combined).
- **Contributions** are computed as
  $w_g\,(I_{g,t}-I_{g,t-12}) \,/\, \sum_h w_h I_{h,t-12}$, where last year's index
  is recovered from the published year-on-year rate. They reproduce the headline
  rate up to rounding in the published indices.
- **State map.** State/UT rates are the CPI (General) year-on-year change for the
  Combined sector, from the same API. Boundaries are as published by the
  Government of India (Survey of India depiction — Jammu & Kashmir and Ladakh
  shown in full), taken from the district GeoJSON of
  [udit-001/india-maps-data](https://github.com/udit-001/india-maps-data)
  and dissolved to state level.
<!-- - **Caching.** API pulls are cached in `data_cache/`; the current year refreshes -->
<!--   after 12 hours, history monthly. Delete the folder to force a full re-download. -->
<!--   New CPI data is normally released at 4 pm IST around the 12th of each month. -->
<!-- - **Refresh** with `quarto render inflation-tracker.qmd` (or the Render button in -->
<!--   RStudio). -->
